---
title: "Analysis of Diminishing Differences Between Naive and Memory in CD4 ChIP-Seq dataset"
author: "Ryan C. Thompson"
date: '`r gsub("\\s+", " ", format(Sys.time(), "%B %e, %Y"))`'
output:
  html_notebook: default
  html_document: default
subtitle: '`r paste0("For histone mark ", params$histone_mark)`'
params:
  histone_mark:
    value: H3K27me3
---

# Preliminary Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, retina=2, cache=TRUE, autodep=TRUE,
                      cache.extra = list(params=params),
                      # https://github.com/yihui/knitr/issues/572
                      cache.lazy=FALSE,
                      fig.height=8, fig.width=8,
                      cache.path = paste0(
                          here::here("cache", "chipseq-diminish", params$histone_mark),
                          .Platform$file.sep))
```

First we load the necessary libraries, along with a set of utility functions.

```{r load_packages, message=FALSE, cache=FALSE}
library(here)
library(stringr)
library(glue)
library(magrittr)
library(openxlsx)
library(SummarizedExperiment)
library(dplyr)
library(edgeR)
library(limma)
library(csaw)
library(sva)
library(ggplot2)
library(scales)
library(GGally)
library(ggalt)
library(ggthemes)
library(splines)
library(reshape2)
library(assertthat)
library(ggfortify)
library(broom)
library(ks)
library(RColorBrewer)

library(BSgenome.Hsapiens.UCSC.hg38)

library(doParallel)
ncores <- getOption("mc.cores", default=parallel::detectCores(logical = FALSE))
options(mc.cores=ncores)
registerDoParallel(cores=ncores)
library(BiocParallel)
register(DoparParam())

options(future.globals.maxSize=4 * 1024^3)
library(future)
plan(multicore)

source(here("scripts", "utilities.R"))

# Required in order to use DGEList objects with future
length.DGEList <- function(x) {
    length(unclass(x))
}
```

We also set the random seed for reproducibility:

```{r set_random_seed}
set.seed(1986)
```

# Data Loading 

We start by loading the already-completed differential modification analysis data.

```{r load_data}
load.filtered(here('saved_data', 'ChIP-seq',
                   glue('{params$histone_mark}-diffmod.rda')),
              exclude="params")
```

# Number of DE genes as a function of time

We begin with a simple analysis by looking at the number of genes significantly differentially expressed between naive and memory at each time point. If differences between the two cell types are diminishing over time, these numbers should show a downward trend. To make sure the number of significant genes is comparable across contrasts, we do a global p-value adjustment using all p-values from all contrasts before thresholding on the resulting FDR values.

```{r num_de_analysis}
nvm.contrasts <- c("NvM.D0", "NvM.D1", "NvM.D5", "NvM.D14")
all.pvals <- peak.results.tables[nvm.contrasts] %>% sapply(. %$% PValue)
all.padj <- all.pvals
all.padj[] %<>% p.adjust(method="BH")
num.sig.table <-
    data.frame(Contrast=colnames(all.padj) %>% fct_inorder,
               FDR05=colSums(all.padj <= 0.05),
               FDR10=colSums(all.padj <= 0.1),
               FDR20=colSums(all.padj <= 0.2))
print(num.sig.table)
num.sig.table %>% gather(Measure, Count, -Contrast) %>% ggplot() + 
    aes(x=Contrast, y=Count, color=Measure, group=Measure) + geom_point() + geom_line()
```

This analysis does not conclusively support a hypothesis of diminishing differences, especially with the small numbers of genes involved at all time points.

# LogFC variance as a function of time

Instead of looking at significant genes, we can also look at the spread of logFC values. To do this, we need to return to the individual window results, since the logFC can be different for each window within a peak. We look at the standard deviation of logFC values at each time point, and we also use a non-parametric test for equality of variances to assess whether there is a significant difference in variance between each pair of successive time points.

```{r logFC_spread}
all.logFC <- window.results.tables[nvm.contrasts] %>% sapply(. %$% logFC) %>%
    as_tibble %>% gather(Contrast, logFC) %>%
    mutate(Contrast=fct_inorder(Contrast))
ggplot(all.logFC) + aes(x=Contrast, y=logFC) + geom_boxplot() + ylab("Window logFC")
all.logFC %>% group_by(Contrast) %>% do(as_tibble(cbind(rbind(summary(.$logFC)), `Std. Dev.`=sd(.$logFC))))
day.comparisons <- list(
    Overall = nvm.contrasts,
    D0vD1 = nvm.contrasts[1:2],
    D1vD5 = nvm.contrasts[2:3],
    D5vD14 = nvm.contrasts[3:4])
lapply(day.comparisons, function(ct) {
    all.logFC %>% filter(Contrast %in% ct) %>%
        fligner.test(logFC ~ Contrast, data=.) %>%
        glance
}) %>% do.call(what=rbind) %>% mutate(Contrast=rownames(.)) %>% select(Contrast, everything())
```

In this case, day 14 actually has the largest logFC variance of any time point. The hypothesis of diminishing differences is not supported for this histone mark.
