---
title: "Differential Modification Analysis of CD4 ChIP-Seq Promoter Dataset"
author: "Ryan C. Thompson"
date: '`r gsub("\\s+", " ", format(Sys.time(), "%B %e, %Y"))`'
output:
    html_document: default
    html_notebook: default
subtitle: '`r glue::glue_data(params, "For histone mark {histone_mark} using {transcriptome} annotation on {genome} genome")`'
params:
    genome:
        value: hg38.analysisSet
    transcriptome:
        value: ensembl.85
    histone_mark:
        value: H3K27me3
    promoter_radius:
        value: 2.5kbp
    fragment_length:
        value: 147bp
---

# Preliminary Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, retina=2, cache=TRUE, autodep=TRUE,
                      cache.extra = list(params=params),
                      # https://github.com/yihui/knitr/issues/572
                      cache.lazy=FALSE,
                      fig.height=8, fig.width=8,
                      cache.path = paste0(
                          here::here("cache", "chipseq-promoter-diffmod", params$histone_mark),
                          .Platform$file.sep))
```

First we load the necessary libraries, along with a set of utility functions.

```{r load_packages, message=FALSE, cache=FALSE}
library(stringr)
library(glue)
library(magrittr)
library(openxlsx)
library(SummarizedExperiment)
library(dplyr)
library(edgeR)
library(limma)
library(csaw)
library(sva)
library(ggplot2)
library(scales)
library(GGally)
library(ggalt)
library(ggthemes)
library(splines)
library(reshape2)
library(assertthat)
library(ggfortify)
library(broom)
library(ks)
library(RColorBrewer)
library(here)

library(BSgenome.Hsapiens.UCSC.hg38)

library(doParallel)
ncores <- getOption("mc.cores", default=parallel::detectCores(logical = FALSE))
options(mc.cores=ncores)
registerDoParallel(cores=ncores)
library(BiocParallel)
register(DoparParam())

options(future.globals.maxSize=4 * 1024^3)
library(future)
plan(multicore)

source(here("scripts/utilities.R"))

# Required in order to use DGEList objects with future
length.DGEList <- function(x) {
    length(unclass(x))
}
```

We also set the random seed for reproducibility:

```{r set_random_seed}
set.seed(1986)
```

# Data Loading and Preprocessing

First we load the consensus peaks called from the reads pooled from all samples. This consensus peak set is not biased toward or against any sample or condition, and therefore the peak significance is expected to be independent of any differential binding in that peak.

```{r load_peaks}
peakfile <- here(
    "peak_calls", "epic_hg38.analysisSet",
    str_c(params$histone_mark, "_condition.ALL_donor.ALL"),
    "peaks_noBL_IDR.narrowPeak")
allpeaks <- {
    read.narrowPeak(peakfile) %>% as("GRanges") %>%
        assign_into(seqinfo(.), seqinfo(BSgenome.Hsapiens.UCSC.hg38)[seqlevels(.)]) %>%
        setNames(.$name)
}
```

Now we'll load the ChIP-seq read count data set from RDS files containing SummarizedExperiment objects, and modify them to use the sample names as column names. We also ensure that the column order is identical between the two objects. Lastly, we filter out any promoters with fewer than one count per sample. This is a very mild filtering criterion, but it sometimes eliminates many promoters, greatly easing the subsequent computational burden of computing the *real* filtering threshold.

```{r load_counts}
sexpfile <-
    here("saved_data", glue_data(params, "promoter-counts_{genome}_{transcriptome}_{promoter_radius}-radius_147bp-reads_{histone_mark}.RDS"))
full.sexp <- readRDS(sexpfile)
colnames(full.sexp) <- colData(full.sexp)$SampleName
sexp <- full.sexp %>% .[rowSums(assay(.)) >= ncol(.),]
```

We extract the sample metadata from the SummarizedExperiment. We set all factors to use a sum-to-zero variant of the treatment-contrast coding, which will ease the subtraction of batch effects later.

```{r extract_samplemeta}
sample.table <- colData(sexp) %>%
    as.data.frame %>% autoFactorize %>%
    mutate(days_after_activation=time_point %>% str_extract("\\d+$") %>% as.numeric(),
           time_point=factor(days_after_activation) %>% `levels<-`(glue("D{levels(.)}")),
           group=interaction(cell_type, time_point, sep="")) %>%
    autoFactorize %>%
    set_rownames(colnames(sexp))
for (i in names(sample.table)) {
    if (is.factor(sample.table[[i]]) && nlevels(sample.table[[i]]) > 1) {
        sample.table[[i]] %<>% C(code_control_named(levels(.)))
    }
}
```

# Peak and Promoter Filtering

We first select all peaks  with an IDR value of 0.2 or less. This is a quite relaxed threshold in order to ensure that even peaks present in only one condition will be included.

# Filtering unbound promoters

We begin by selecting only peaks  with an IDR value of 0.05 or less, and then determine the set of promoters that overlap these peaks.

```{r filter_peaks}
idr.threshold <- 0.05
genome.size <- seqlengths(seqinfo(allpeaks)) %>% as.numeric %>% sum
# IDR is encoded in the qValue field of the narrowPeak file
peaks <- allpeaks[allpeaks$qValue >= -log10(idr.threshold)]
pct.genome.covered <- width(peaks) %>% sum %>% divide_by(genome.size) %>% multiply_by(100)
peak.overlap <- overlapsAny(sexp, peaks)
pct.promoters.covered <- mean(peak.overlap) * 100
message(glue("Selected {length(peaks)} peaks at an IDR threshold of {format(idr.threshold, digits=3)}, with an average width of {round(mean(width(peaks)))}, covering a total of {format(pct.genome.covered, digits=3)}%% of the genome and {format(pct.promoters.covered, digits=3)}% of promoters."))
```

```{r filter_promoters}
a <- aveLogCPM(asDGEList(sexp), prior.count = 2)
adata <- data.frame(logCPM=a, PeakOverlap=peak.overlap)
threshold.q <- 0.05
logcpm.threshold <- adata %>% filter(PeakOverlap) %$% quantile(logCPM, threshold.q)
count.threshold <- 2^logcpm.threshold * mean(colData(sexp)$totals) / 1e6
message(glue("
    Filter theshold at {threshold.q} quantile of peak-overlapping promoters is {format(count.threshold, digits=3)} reads, a logCPM of {format(logcpm.threshold, digits=3)}. \\
    This threshold keeps {format(100 * mean(a >= logcpm.threshold), digits=3)}% of promoters.
"))
sexp <- sexp[a >= logcpm.threshold,]
```



Now we create a DGEList from the counts.

```{r prepare_dgelist}
## Extract gene metadata and colapse lists
all.promoter.meta <- rowRanges(sexp) %>% as.data.frame %>%
    select(-width, -strand) %>% rename(chr=seqnames)
# Convert list columns to character vectors
all.promoter.meta[] %<>% lapply(function(x) if (is.list(x)) sapply(x, str_c, collapse=",") else x)
dge <- asDGEList(sexp) %>%
    assign_into(.$offset, NULL) %>%
    assign_into(.$genes, all.promoter.meta) %>%
    set_rownames(rownames(all.promoter.meta))
```

Then we select all promoters with an average read count of at least 5.

## Model fitting

Now we are ready to fit a model to the data. We begin by building the design matrix, inclulding coefficients for the interaction of cell type and time point (group). We also construct an equivalent design matrix that includes an intercept for use in methods that don't work on intercept-free designs.

```{r build_design_matrix}
design <- model.matrix(~0 + group, sample.table, strip.prefixes = TRUE)
design.int <- model.matrix(~1 + group, sample.table)
colnames(design)
```

Instead of incorporating all the batch effects we think might be important into the design matrix, we let SVA infer the important confounding factors from the data, and add those to the design matrix.

```{r sva}
dge %<>% calcNormFactors()
logcpm <- cpm(dge, log=TRUE, prior.count=1)
svobj <- sva(logcpm, design.int)
svmat <- add.numbered.colnames(svobj$sv, "SV")
design.sv <- cbind(design, svmat)
```

We plot the surrogate variables against known variables and also test for significant correlations between them.

```{r test_sv_corr}
sv.cor.tests <- expand.grid(SV=colnames(svmat),
                            Var=c("cell_type", "time_point", "group", "donor_id", "totals")) %>%
    select(Var, SV) %>%
    group_by_all() %>% 
    do(mod=list(lm(as.formula(glue("{.$SV} ~ {.$Var}")), data=cbind(sample.table, svmat)))) %>%
    mutate(aov=lapply(mod, anova)) %>%
    ungroup() %>%
    mutate(PValue=vapply(.$aov, . %$% {`Pr(>F)`[1]}, numeric(1)),
           FDR=p.adjust(PValue, method="BH")) %>%
    select(-mod, -aov)
sv.cor.tests %>% arrange(PValue)
ggduo.dataXY(sample.table[c("cell_type", "time_point", "donor_id", "totals")], svmat,
             extraData=sample.table["group"],
             types=list(comboVertical="dot_no_facet"),
             mapping=aes(color=group, shape=donor_id))
```
We now proceed to computing normalization factors, estimating dispersions, and fitting the quasi-likelihood GLM.

```{r fit_model}
dge %<>% estimateDisp(design.sv, robust=TRUE)
plotBCV(dge)
fit <- glmQLFit(dge, design.sv, robust=TRUE)
plotQLDisp(fit)
```

Next, we define the differential expression tests we wish to perform as contrasts. Each contrast is an arithmetic expression in terms of the model coefficients.

```{r define_contrasts}
celltypes <- levels(sample.table$cell_type)
all.timepoints <- levels(sample.table$time_point)
nonzero.timepoints <- setdiff(all.timepoints, "D0")

timepoint.anova.tests <- setNames(llply(celltypes, function(ct) {
    setNames(glue("{ct}{nonzero.timepoints} - {ct}D0"),
             glue("{ct}.D0v{nonzero.timepoints}"))
}), nm=str_c(celltypes, ".AllT"))
timepoint.single.tests <- as.list(unlist(unname(timepoint.anova.tests)))
celltype.singlet.tests <-
    as.list(setNames(glue("Memory{all.timepoints} - Naive{all.timepoints}"),
                     glue("NvM.{all.timepoints}")))
celltype.allt.test <- list(NvM.AllT=unlist(celltype.singlet.tests))
factorial.singlet.tests <-
    as.list(setNames(glue("(Memory{nonzero.timepoints} - MemoryD0) - (Naive{nonzero.timepoints} - NaiveD0)"),
                     glue("Fac.{nonzero.timepoints}")))
factorial.allt.test <- list(Fac.AllT=unlist(factorial.singlet.tests))
mi.vs.nf.test <- list(MD0vND14="MemoryD0 - NaiveD14")
sv.test <- list(SV=colnames(svmat))
alltests <- c(timepoint.anova.tests, timepoint.single.tests,
              celltype.allt.test, celltype.singlet.tests,
              factorial.allt.test, factorial.singlet.tests,
              mi.vs.nf.test,
              sv.test
              ) %>% lapply(as_glue)
print(alltests)
```

We now perform the differential modification tests for each contrast or set of contrasts. For a single contrast, the test is analogous to a t-test. For a multi-contrast test, the global null hypothesis that all contrasts are equal to zero is used, analogous to an F-test.

```{r test_contrasts}
results.tables <- bplapply(alltests, function(ct) {
    ctmat <- makeContrasts(contrasts = ct, levels=design.sv) %>% set_colnames(names(ct))
    ctest <- glmQLFTest(fit, contrast = ctmat)
    topTags(ctest, n=Inf, sort.by="none") %>% as.data.frame %>% add.qvalue
})
```

```{r pval_hist}
p <- bplapply(names(results.tables), function(testname) {
    pvals <- results.tables[[testname]]$PValue
    pi0 <- pi0est(pvals)$pi0
    p <- plotpvals(pvals, ptn = pi0) +
        labs(title=glue("P-value histogram for {testname}"),
             subtitle=glue("Est. Non-Null Prop.: {(format((1 - pi0) * 100, digits=3))}%%"))
})
ggprint(p)
```

```{r num_signif}
results.tables %>% lapply(function(rt) {
    summarize(rt,
              NumSignif=FDR %>% is_weakly_less_than(0.1) %>% sum,
              EstNonNull=floor(length(PValue) * (1 - propTrueNull(PValue))))
}) %>% 
    do.call(what=rbind) %>% 
    mutate(Test=rownames(.)) %>%
    select(Test, everything())
```

```{r save_results}
# Take all genes significant at 10% FDR or top 100 genes, whichever is more
filtered.results.tables <- lapply(results.tables, . %>% arrange(PValue) %>% filter(FDR <= 0.1 | seq_len(nrow(.)) <= 100))
dir.create(here("saved_data", "ChIP-seq"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results", "ChIP-seq"), recursive = TRUE, showWarnings = FALSE)
saveRDS(results.tables,
    here('saved_data', 'ChIP-seq',
        glue_data(params, '{genome}_{transcriptome}_{histone_mark}-{promoter_radius}-promoter-diffmod.RDS')))
save.image(here('saved_data', 'ChIP-seq',
    glue_data(params, '{genome}_{transcriptome}_{histone_mark}-{promoter_radius}-promoter-diffmod.rda')))
write.xlsx(filtered.results.tables,
    here("results", "ChIP-seq",
        glue_data(params, '{genome}_{transcriptome}_{histone_mark}-{promoter_radius}-promoter-diffmod.xlsx')))
```
