---
title: "MOFA Model Fitting for RNA-seq and Promoter Histone ChIP-seq data"
author: "Ryan C. Thompson"
date: '`r gsub("\\s+", " ", format(Sys.time(), "%B %e, %Y"))`'
output:
    html_document: default
    html_notebook: default
params:
    mofa_runs: 4
    genome:
        value: hg38.analysisSet
    transcriptome:
        value: ensembl.85
    rna_dataset:
        value: "shoal_hg38.analysisSet"
    promoter_datasets:
        value: !r c(H3K4me2 = "1kbp-radius_147bp-reads_H3K4me2", 
                    H3K4me3 = "1kbp-radius_147bp-reads_H3K4me3", 
                    H3K27me3 = "2.5kbp-radius_147bp-reads_H3K27me3")
    peak_datasets:
        value: !r c(H3K4me2 = "H3K4me2_147bp-reads", 
                    H3K4me3 = "H3K4me3_147bp-reads", 
                    H3K27me3 = "H3K27me3_147bp-reads")
    window_datasets:
        value: !r c(H3K4me2 = "500bp-windows_147bp-reads_H3K4me2", 
                    H3K4me3 = "500bp-windows_147bp-reads_H3K4me3", 
                    H3K27me3 = "500bp-windows_147bp-reads_H3K27me3")
---

# Preliminary setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, retina=2, cache=TRUE, autodep=TRUE,
                      cache.extra = list(params=params),
                      # https://github.com/yihui/knitr/issues/572
                      cache.lazy=FALSE,
                      fig.height=8, fig.width=8,
                      cache.path = paste0(
                          here::here("cache", "promoter-mofa-run"),
                          .Platform$file.sep))
```

First we load the necessary libraries, along with a set of utility functions.

```{r load_packages, message=FALSE, cache=FALSE}
library(stringr)
library(glue)
library(magrittr)
library(openxlsx)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(edgeR)
library(limma)
library(DESeq2)
library(csaw)
library(sva)
library(MOFAtools)
library(ggplot2)
library(scales)
library(GGally)
library(ggalt)
library(ggthemes)
library(splines)
library(reshape2)
library(assertthat)
library(ggfortify)
library(broom)
library(ks)
library(RColorBrewer)
library(here)

library(BSgenome.Hsapiens.UCSC.hg38)

library(doParallel)
ncores <- getOption("mc.cores", default=parallel::detectCores(logical = FALSE))
options(mc.cores=ncores)
registerDoParallel(cores=ncores)
library(BiocParallel)
register(SerialParam())
register(DoparParam())

options(future.globals.maxSize=4 * 1024^3)
library(future)
plan(multicore)

# Export environment variables to make NumPy respect the desired level of 
# parallelism. (Different variables are required for different libraries.)
Sys.setenv(MKL_NUM_THREADS = ncores,
           NUMEXPR_NUM_THREADS = ncores,
           OMP_NUM_THREADS = ncores)

source(here("scripts/utilities.R"))

# Required in order to use DGEList objects with future
length.DGEList <- function(x) {
    length(unclass(x))
}
```

# Data loading

First we load all the count data sets that we will be using.

```{r load_counts}
promoter.sexps %<-% {
    bplapply(here("saved_data",
                  glue_data(params, "promoter-counts_{genome}_{transcriptome}_{promoter_datasets}.RDS")),
             readRDS) %>% 
        setNames(names(params$promoter_datasets))
}
rnaseq.sexp %<-% {
    readRDS(here("saved_data", glue_data(params, "SummarizedExperiment_rnaseq_{rna_dataset}_{transcriptome}.RDS")))  %>%
    set_colnames(colData(.)$SRA_run)
}
for (i in names(promoter.sexps)) {
    promoter.sexps[[i]] %<>%
        set_colnames(colData(.)$SRA_run)
    colData(promoter.sexps[[i]]) %<>%
        transform(time_point = str_replace(time_point, "Day", "D")) %>%
        transform(donor_id = str_replace(donor_id, "^Dn*", "Dn")) %>%
        transform(PrimarySample = glue("{cell_type}-{time_point}-{donor_id}"))
}
colData(rnaseq.sexp) %<>% 
    transform(time_point = glue("D{days_after_activation}")) %>%
    transform(donor_id = str_replace(donor_id, "^Dn*", "Dn")) %>%
    transform(PrimarySample = glue("{cell_type}-{time_point}-{donor_id}"))

# Fix up stranded RNA-seq counts, if needed
libtype.assayNames <- c(SF="sense.counts", SR="antisense.counts")
if (all(libtype.assayNames %in% assayNames(rnaseq.sexp))) {
    message("Selecting stranded counts for each sample")
    sample.table %<>% mutate(count_type=libtype.assayNames[libType])
    assay(rnaseq.sexp, "unstranded.counts") <- assay(rnaseq.sexp, "counts")
    assay(rnaseq.sexp, "counts") <- lapply(seq_len(nrow(sample.table)), function(i) {
        message("Using ", sample.table[i,]$count_type, " for ", colnames(rnaseq.sexp)[i])
        assay(rnaseq.sexp, sample.table[i,]$count_type %>% as.character)[,i]
    }) %>% do.call(what=cbind)
    total.counts <- sexp %>% assays %>% sapply(colSums) %>% data.frame %>%
        mutate(SampleName=row.names(.)) %>%
        inner_join(sample.table, by="SampleName")
    total.counts %$% invisible(assert_that(all(counts == pmax(sense.counts, antisense.counts))))
}

sexps <- c(list(RNA=rnaseq.sexp), promoter.sexps)
```

# Normalization and filtering

## Abundance filtering

The first filter is an abundance filter, which just re-uses the filtering criteria previously defined in other reports.

```{r filter_abundance}
presence.thresholds <- list(RNA=-1, H3K4me2=0.407, H3K4me3=1.29, H3K27me3=0.993)
#presence.thresholds <- list(RNA=-1, H3K4me2=1+0.407, H3K4me3=1+1.29, H3K27me3=0.5+0.993)
# Filter RNA-seq at average logCPM >= -1
for (i in names(presence.thresholds)) {
    suppressWarnings(present <- aveLogCPMWithOffset(asDGEList(sexps[[i]])) >= presence.thresholds[[i]])
    num.features <- nrow(sexps[[i]])
    num.kept <- sum(present)
    percent.kept <- num.kept / num.features * 100
    message(glue("For data type {i}, keeping {num.kept} features out of {num.features} ({format(percent.kept, digits=3)}%) with aveLogCPM >= {presence.thresholds[[i]]}"))
    sexps[[i]] %<>% .[present,]
    rm(present)
}
```

## Calculating normalization factors

We need to normalize and filter all data sets before feeding them to MOFA. We begin by computing normalization factors. (potentially with offsets derived from effective gene lengths for RNA-seq.)

```{r normalize_data}
for (i in names(sexps)) {
    if (! "totals" %in% colnames(colData(sexps[[i]]))) {
        colData(sexps[[i]])$totals <- colSums(assay(sexps[[i]], "counts"))
    }
    colData(sexps[[i]])$norm.factors <- calcNormFactors(asDGEList(sexps[[i]]))$samples$norm.factors
    
    if ("length" %in% assayNames(sexps[[i]])) {
        normMat <- assay(sexps[[i]], "length") %>% divide_by(exp(rowMeans(log(.))))
        normCounts <- assay(sexps[[i]], "counts")/normMat
        lib.offsets <- log(calcNormFactors(normCounts)) + log(colSums(normCounts))
        assay(sexps[[i]], "offset") <- t(t(log(normMat)) + lib.offsets)
    }
}
```

## Outlier sample filtering

Next, we identify possible outlier samples in the data. For our purposes, these are defined as samples for which the percentage of zero counts is 3 standard deviations below the mean for all samples. This criterion was determined through exploratory data analysis. We drop these samples before running MOFA.

```{r filter_outlier_samples}
outlier.samples <- sexps %>% 
    lapply(. %>% 
               set_colnames(colData(.)$PrimarySample) %>%
               assay("counts") %>%
               is_greater_than(0) %>% 
               colMeans %>%
               .[. < mean(.) - sd(.) * 3] %>% 
               names)
for (i in names(sexps))  {
    outliers <- sexps[[i]]$PrimarySample %in% outlier.samples[[i]]
    if (any(outliers)) {
        message(glue("Removing out {sum(outliers)} outlier sample{ifelse(sum(outliers) == 1, '', 's')} from {i} data."))
    }
    sexps[[i]] %<>% .[,!outliers]
}
```

## Variance filtering

The next filter is a variance filter, selecting N genes/peaks from each data set with the largest dispersions. The value of N for each data set is determined based on the number of significantly differentially abundant features identified in previous analyses.

```{r filter_variance}
num.keep <- list(RNA=10000, H3K4me2=20000, H3K4me3=10000, H3K27me3=20000)
#num.keep <- list(RNA=10000, H3K4me2=10000, H3K4me3=10000, H3K27me3=10000)
for (i in names(num.keep)) {
    if (num.keep[[i]] >= nrow(sexps[[i]])) {
        message(glue("For data type {i}, no variance filter is needed; keeping all {nrow(sexps[[i]])} features"))
    } else {
        assert_that(num.keep[[i]] <= nrow(sexps[[i]]))
        d <- asDGEList(sexps[[i]])
        design <- matrix(1, nrow=ncol(d), ncol=1)
        d <- estimateDisp(d, design, prior.df = 0)
        plotBCV(d, main = glue("BCV plot for {i}"))
        ostat <- nrow(d) - num.keep[[i]] + 1
        disp.threshold <- d$tagwise.dispersion %>% sort(partial=ostat) %>% .[ostat]
        message(glue("For data type {i}, keeping the top {num.keep[[i]]} features with the highest dispersions out of {nrow(d)}"))
        keep <- d$tagwise.dispersion >= disp.threshold
        assert_that(sum(keep) == num.keep[[i]])
        sexps[[i]] %<>% .[keep,]
    }
}
```

## Data Transformation

Now we transform all the filtered datasets to log2 CPM, using a prior count of 2, since that is what plotMDS uses, and MOFA is another factor analysis method similar to MDS or PCA.

```{r logcpm_transform}
logcpmlist <- lapply(sexps, . %>% asDGEList %>% cpmWithOffset(log=TRUE, prior.count=2))
```

## Preparing the data for MOFA

Finally, we are ready to combine all filtered, transformed datasets into a MultiAssayExperiment.

```{r build_MAE}
make.samplemap <- function(explist, primary_colname="primary") {
    maps <- lapply(explist, . %>% {data.frame(primary=colData(.)[[primary_colname]], colname=colnames(.),
                                              stringsAsFactors = FALSE)})
    x <- listToMap(maps)
}
biosample.meta <- colData(rnaseq.sexp)[c("cell_type", "activated", "time_point", "days_after_activation", "donor_id")] %>%
    transform(
        time_point = factor(time_point, levels=unique(time_point[order(days_after_activation)])),
        donor_id = factor(donor_id),
        PrimarySample = glue("{cell_type}-{time_point}-{donor_id}")) %>%
    set_rownames(.$PrimarySample)
smap <- make.samplemap(sexps, "PrimarySample")
# Since MOFA is related to MDS, we use the same prior count as plotMDS
mae <- MultiAssayExperiment(experiments = logcpmlist, sampleMap = smap, colData = biosample.meta)
```

We run MOFA several times with different random seeds so that we can verify that it consistently converges to the same result. Then we run it once with a tighter tolerance bound to obtain the final model.

```{r prepare_mofa_options}
mofa <- createMOFAobject(mae)
ModelOptions <- getDefaultModelOpts(mofa)
TrainOptions.final <- getDefaultTrainOpts() %>% transform(maxiter=30000)
# Looser tolerance bound for exploration
TrainOptions.explore <- getDefaultTrainOpts() %>% transform(tolerance=0.1)

tmpd <- tempfile(tmpdir = tempdir(), pattern = "mofatemp")
DirOptions.final <- list(
    dataDir = tempfile(tmpdir = tmpd, pattern = "mofarun_"),
    outFile = here("saved_data", "mofa", glue_data(params, "mofa-model_{genome}_{transcriptome}_rna+promoter.hdf5")),
    rdsFile = here("saved_data", "mofa", glue_data(params, "mofa-model_{genome}_{transcriptome}_rna+promoter.RDS"))
)
DirOptions.exploreList <- lapply(seq_len(params$mofa_runs), function(i) {
    list(
        dataDir = tempfile(tmpdir = tmpd, pattern = str_c("mofarun", i, "_")),
        outFile = here("saved_data", "mofa", glue_data(params, "mofa-model_{genome}_{transcriptome}_rna+promoter_explore{i}.hdf5")),
        rdsFile = here("saved_data", "mofa", glue_data(params, "mofa-model_{genome}_{transcriptome}_rna+promoter_explore{i}.RDS"))
    )
})
final.random.seed <- 1986
explore.random.seeds <- final.random.seed + seq_len(params$mofa_runs)
```

# Model fitting

```{r fit_explore_models}
# MOFA is already parallelized, so we run sequentially
mofa.explore <- list()
for (i in seq_len(params$mofa_runs)) {
    DirOpt <- DirOptions.exploreList[[i]]
    seed <- explore.random.seeds[i]
    dir.create(DirOpt$dataDir, recursive=TRUE)
    mofa.explore[[i]] <- mofa %>% 
        prepareMOFA(DirOptions = DirOpt, ModelOptions = ModelOptions, TrainOptions = TrainOptions.explore) %>%
        runMOFA(DirOptions = DirOpt, seed=seed)
    saveRDS(mofa.explore[[i]], DirOpt$rdsFile)
    ## TODO: saveRDS
}
unlink(tmpd, recursive=TRUE)
```

```{r fit_final_model}
dir.create(DirOptions.final$dataDir, recursive=TRUE)
mofa.final <- mofa %>%
    prepareMOFA(DirOptions = DirOptions.final, ModelOptions = ModelOptions, TrainOptions = TrainOptions.final) %>%
    runMOFA(DirOptions = DirOptions.final, seed=final.random.seed)
saveRDS(mofa.final, DirOptions.final$rdsFile)
unlink(tmpd, recursive=TRUE)
#mofa.final <- mofa.explore[[1]]
## TODO: saveRDS
```

# Basic model QC

After the run finishes. We produce a few basic QC plots. First, we plot the variance explained by each factor in each model.

```{r plot_r2, warning = FALSE}
r2list <- lapply(mofa.explore, calculateVarianceExplained)
r2.final <- calculateVarianceExplained(mofa.final)
```

We also make a plot comparing the factors between multiple models, to verify that each model is discovering roughly the same set of factors.

```{r compare_models}
#invisible(compareModels(mofa.explore))
```

Since all the models seem to be discovering the same factors, we can use any model we choose. For the first model, we make a "bee swarm plot" of each factor:

```{r bee_swarm_plots}
# p <- list(
#     TimePoint=plotFactorBeeswarm(
#         mofa.final, 
#         factors = seq_len(ncol(getFactors(mofa.final, include_intercept = FALSE))), 
#         color_by = "time_point") + 
#         aes(x=0) + xlab("") + ylab("Factor Value") + 
#         facet_wrap(~factor, scales="free") + 
#         scale_x_continuous(breaks=NULL),
#     CellType=plotFactorBeeswarm(
#         mofa.final, 
#         factors = seq_len(ncol(getFactors(mofa.final, include_intercept = FALSE))), 
#         color_by = "cell_type") + 
#         aes(x=0) + xlab("") + ylab("Factor Value") + 
#         facet_wrap(~factor, scales="free") + 
#         scale_x_continuous(breaks=NULL),
#     Donor=plotFactorBeeswarm(
#         mofa.final, 
#         factors = seq_len(ncol(getFactors(mofa.final, include_intercept = FALSE))), 
#         color_by = "donor_id") + 
#         aes(x=0) + xlab("") + ylab("Factor Value") + 
#         facet_wrap(~factor, scales="free") + 
#         scale_x_continuous(breaks=NULL))
# ggprint(p)
```
