---
title: "Differential expression analysis of CD4 RNA-Seq Dataset"
author: "Ryan C. Thompson"
date: '`r gsub("[[:space:]]+", " ", format(Sys.time(), "%B %e, %Y"))`'
output: html_notebook
subtitle: '`r paste0("Using Dataset ", params$dataset)`'
params:
    genome:
        value: hg38.analysisSet
    transcriptome:
        value: ensembl.85
    quant_method:
        value: shoal
---

# Preliminary Setup

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, retina = 2, cache = TRUE, autodep = TRUE,
                      cache.extra = list(params = params),
                      fig.height = 8, fig.width = 8,
                      cache.path = paste0(
                          here::here("cache", "rnaseq-diffexp", params$dataset),
                          .Platform$file.sep))
```

First we load the necessary libraries, along with a set of utility functions.

```{r load_packages, message = FALSE, cache = FALSE}
library(here)
library(stringr)
library(limma)
library(rctutils)

library(BiocParallel)
library(future)
options(future.globals.maxSize = 4 * 1024^3)
use_futures("multicore")
```

We also set the random seed for reproducibility:

```{r set_random_seed}
set.seed(1986)
```

# Data Loading and Preprocessing

We begin by loading the saved data from the RNA-seq differential expression tests. We will need these fitted models to run the gene set tests.

```{r load_diffexp_results}
diffexp <- load_in_new_env(here('saved_data', 'RNA-seq', glue_data(params, '{quant_method}_{genome}_{transcriptome}-diffexp.rda')))
design <- diffexp$design
alltests <- diffexp$alltests
elist.cb <- diffexp$elist.cb
```

We also need to load the gene sets representing TFBS overlap with gene promoters.

## Loading the TFBS overlap sets

We load the TFBS overlap data for all peaks. For RNA-seq genes, we say that a gene has a TFBS if any of its peaks overlap one.

```{r load_tfbs_overlap}
overlap_sets <- readRDS(here("saved_data", glue_data(params, "peak-tfbs-overlap_{genome}.RDS")))
promoter_overlap_sets <- readRDS(here("saved_data", glue_data(params, "promoter-tfbs-overlap_{genome}_{transcriptome}.RDS")))
overlap_sets$RNA <- bplapply(names(promoter_overlap_sets[[1]]), function(tf) {
    lapply(overlap_sets, . %>% .[[tf]] %>% str_replace("-P[0-9]+$", "")) %>%
        unlist %>% sort %>% unique
}) %>% set_names(names(overlap_sets[[1]]))
```

# CAMERA

We now run the CAMERA gene set test on all the TFBS overlap sets. We run CAMERA on each single-contrast differential expression test that was run.

```{r run_camera}
single.contrast.tests <- alltests %>% .[lengths(.) == 1]
camera.results <- bplapply(single.contrast.tests, function(ct) {
    cmat <- makeContrasts(contrasts = ct, levels = design)
    camera(fit)
})
```
