---
title: "TFBS Overlap Determination for Promoters and Peaks"
author: "Ryan C. Thompson"
date: '`r stringr::str_replace_all(format(Sys.time(), "%B %e, %Y"), "[[:space:]]+", " ")`'
output:
    html_document: default
    html_notebook: default
params:
    genome:
        value: hg38.analysisSet
    transcriptome:
        value: ensembl.85
    promoter_datasets:
        value: !r c(H3K4me2 = "1kbp-radius_147bp-reads_H3K4me2",
                    H3K4me3 = "1kbp-radius_147bp-reads_H3K4me3",
                    H3K27me3 = "2.5kbp-radius_147bp-reads_H3K27me3")
    peak_datasets:
        value: !r c(H3K4me2 = "epic_H3K4me2_147bp-reads",
                    H3K4me3 = "epic_H3K4me3_147bp-reads",
                    H3K27me3 = "epic_H3K27me3_147bp-reads")
---

# Preliminary setup

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, retina = 2, cache = TRUE, autodep = TRUE,
                      cache.extra = list(params = params),
                      # https://github.com/yihui/knitr/issues/572
                      cache.lazy = FALSE,
                      fig.height = 8, fig.width = 8,
                      cache.path = paste0(
                          here::here("cache", "tfbs-overlap"),
                          .Platform$file.sep))
```

First we load the necessary libraries, along with a set of utility functions.

```{r load_packages, message = FALSE, cache = FALSE}
library(glue)
library(SummarizedExperiment)
library(here)
library(rctutils)


library(BiocParallel)
library(future)
options(future.globals.maxSize = 4 * 1024^3)
use_futures("multicore")
```

# Data loading

## Loading promoters

First we load all the data sets that we will be using.

```{r load_counts}
promoter_sexps <-
    bplapply(here("saved_data",
                  glue_data(params, "promoter-counts_{genome}_{transcriptome}_{promoter_datasets}.RDS")),
             readRDS) %>%
        setNames(names(params$promoter_datasets))
peak_sexps <-
    bplapply(here("saved_data",
                  glue_data(params, "peak-counts_{genome}_{peak_datasets}.RDS")),
             readRDS) %>%
        setNames(names(params$promoter_datasets))
```


## Loading transcription factor binding sites

```{r load_tfbs_cons}
all_tfbs <- readRDS(here("saved_data", "UCSC_tfbsCons.RDS"))
tfbs <- all_tfbs %>% split(.$name)
```

# Computing overlaps

We define a "feature set" for each transcription factor as the set of promoters that overlap or contain at least one binding site for that transcription factor.

```{r compute_overlap_sets}
promoter_overlap_sets <- lapply(promoter_sexps, function(sexp) {
    bplapply(tfbs, . %>% overlapsAny(query = sexp) %>% {rownames(sexp)[.]})
})
peak_overlap_sets <- lapply(peak_sexps, function(sexp) {
    bplapply(tfbs, . %>% overlapsAny(query = sexp) %>% {rowData(sexp)$name[.]})
})
```

# Save data

```{r save_overlap_sets}
saveRDS(promoter_overlap_sets, here("saved_data", glue_data(params, "promoter-tfbs-overlap_{genome}_{transcriptome}.RDS")))
saveRDS(peak_overlap_sets, here("saved_data", glue_data(params, "peak-tfbs-overlap_{genome}.RDS")))
```

